### 2021-06-11 misc/proofs/coprime-probability.tex
Spacing and punctuation fix
Lattice of arithmetic progressions

If two integers $u,v$ have no prime factors in common (i.e.\ $\gcd(u,v) = 1$), we say they are {\it relatively prime} and write $u\perp v$. For any positive integer $n$, let $q_n$ denote the number of ordered pairs of integers $(u,v)$ such that $1\leq u,v\leq n$and $u\perp v$. Then the probability that any pair of integers $u,v$ lying in the range $[1\ ..\ n]$ are relatively prime is $q_n/n^2$. Our goal is to find out what happens when $u$ and $v$ may be {\it any} positive integers. In other words, we will investigate the behaviour of $q_n/n^2$ as $n\rightarrow \infty$. We begin with the following lemma:

@@

\proof To prove this limit, we will investigate the difference
$$q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 = \sum_{k\geq 1}\mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2.$$
Because $\lfloor n/k \rfloor = 0$ for $k>n$, the first summation on the right-hand side need only range up to $k=n$, and we can reorganise the summations as follows:
$$\eqalignno{
    \sum_{k\geq 1}\mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n \mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k=1}^n \mu(k)\left({n\over k}\right)^2 - \sum_{k>n}\mu(k)\left({n\over k}\right)^2\cr
    q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n \mu(k) \left(\left\lfloor {n\over k} \right\rfloor ^2 - \left({n \over k}\right)^2\right) - \sum_{k>n}\mu(k)\left({n\over k}\right)^2 & ({\oldstyle 1})
}$$
We will want to put asymptotic bounds on these rearranged sums. First, note that for all positive real $x$,
$$x^2 - \lfloor x\rfloor^2 = O(x).\oldno 2$$
For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x$. Next, we use standard formulas to analyse the asymptotic behaviour of the series $\sum_{k>n}(n/k)^2$:
$$\eqalign{
     \sum_{k>n} \left({n\over k}\right)^2 &= \sum_{k\geq 1} \left({n\over k}\right)^2 - \sum_{k=1}^n \left({n\over k}\right)^2 \cr
    &= {n^2\pi^2\over 6} - n^2\left({\pi^2\over 6} -  {1\over n} + O( n^{-2} )\right) \cr
    &= n - O(1) \cr
    &= O(n)
}\oldno 3$$
Since $\mu(k)\leq 1$, we can discard this factor when applying Eqs.\ ({\oldstyle 2}) and ({\oldstyle 3}) to get the asymptotic behaviour of Eq.\ ({\oldstyle 1}); this yields
$$\eqalign{
    q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n O(n/k) - O(n)\cr
    &= O(nH_n) - O(n),
}$$
where $H_n = \sum_{k=1}^n 1/k$ denotes the $n$-th harmonic number. Because both $O(nH_n)$ and $O(n)$ are $o(n^2)$, we can replace these terms by $o(n^2)$ and divide by $n^2$ to arrive at
$${q_n\over n^2} - \sum_{k\geq 1}{\mu(k)\over k^2} = o(1).$$
This means that for any real $\epsilon > 0$, we can find a number $n_0$ such that for all $n \geq n_0$ the difference $|q_n - \sum_{k\geq 1} \mu(k)/k^2| \leq \epsilon$; the sequence $q_n/n^2$ converges exactly as stated by the lemma.\slug

@@

\proof The case $n=1$ is obviously true, so let $n>1$. By the fundamental theorem of arithmetic, $n = p_1^{e_1}p_2^{e_2}\cdots p_k^{e_k}$. Only square-free divisors contribute to the sum $\sum_{d\backslash n}\mu(d)$, so we have
$$\sum_{d\backslash n}\mu(d) = \mu(1) + \sum_{1\leq i\leq k} \mu(p_i) + \sum_{1\leq i<j\leq k} \mu(p_ip_j) + \cdots + \mu(p_1\cdots p_k).$$
Since the magnitude of $\mu(d)$ is 1 (whenever $d$ is square-free) and the sign of $\mu(d)$ depends on the number of primes that make up $d$, this can be re-expressed as the alternating sum of binomial coefficients
$$\sum_{d\backslash n}\mu(d) = {k \choose 0} - {k\choose 1} + {k\choose 2} - \cdots + (-1)^k{k\choose k}.$$
which equals 0. (We will not prove this claim, but one may be convinced of this fact by trying it out on a couple of rows of Pascal's Triangle.)\slug

@@

\proclaim Lemma C.
$$\biggl(\sum_{k\geq 1} {\mu(k) \over k^2}\biggr) \biggl( \sum_{m \geq 1} {1\over m^2}\biggr) = 1$$.

>>>>

If two integers $u,v$ have no prime factors in common (i.e.\ $\gcd(u,v) = 1$), we say they are {\it relatively prime} and write $u\perp v$. For any positive integer $n$, let $q_n$ denote the number of ordered pairs of integers $(u,v)$ such that $1\leq u,v\leq n$ and $u\perp v$. Then the probability that any pair of integers $u,v$ lying in the range $[1\ ..\ n]$ are relatively prime is $q_n/n^2$. Our goal is to find out what happens when $u$ and $v$ may be {\it any} positive integers. In other words, we will investigate the behaviour of $q_n/n^2$ as $n\rightarrow \infty$. We begin with the following lemma:

@@

\proof To prove this limit, we will investigate the difference
$$q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 = \sum_{k\geq 1}\mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2.$$
Because $\lfloor n/k \rfloor = 0$ for $k>n$, the first summation on the right-hand side need only range up to $k=n$, and we can reorganise the summations as follows:
$$\eqalignno{
    \sum_{k\geq 1}\mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n \mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k=1}^n \mu(k)\left({n\over k}\right)^2 - \sum_{k>n}\mu(k)\left({n\over k}\right)^2\cr
    q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n \mu(k) \left(\left\lfloor {n\over k} \right\rfloor ^2 - \left({n \over k}\right)^2\right) - \sum_{k>n}\mu(k)\left({n\over k}\right)^2 & ({\oldstyle 1})
}$$
We will want to put asymptotic bounds on these rearranged sums. First, note that for all positive real $x$,
$$x^2 - \lfloor x\rfloor^2 = O(x),\oldno 2$$
for $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x+1$. Next, we use standard formulas to analyse the asymptotic behaviour of the series $\sum_{k>n}(n/k)^2$:
$$\eqalign{
     \sum_{k>n} \left({n\over k}\right)^2 &= \sum_{k\geq 1} \left({n\over k}\right)^2 - \sum_{k=1}^n \left({n\over k}\right)^2 \cr
    &= {n^2\pi^2\over 6} - n^2\left({\pi^2\over 6} -  {1\over n} + O( n^{-2} )\right) \cr
    &= n - O(1) \cr
    &= O(n)
}\oldno 3$$
Since $\mu(k)\leq 1$, we can discard this factor when applying Eqs.\ ({\oldstyle 2}) and ({\oldstyle 3}) to get the asymptotic behaviour of Eq.\ ({\oldstyle 1}); this yields
$$\eqalign{
    q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n O(n/k) - O(n)\cr
    &= O(nH_n) - O(n),
}$$
where $H_n = \sum_{k=1}^n 1/k$ denotes the $n$-th harmonic number. Because both $O(nH_n)$ and $O(n)$ are $o(n^2)$, we can replace these terms by $o(n^2)$ and divide by $n^2$ to arrive at
$${q_n\over n^2} - \sum_{k\geq 1}{\mu(k)\over k^2} = o(1).$$
This means that for any real $\epsilon > 0$, we can find a number $n_0$ such that for all $n \geq n_0$ the difference $|q_n - \sum_{k\geq 1} \mu(k)/k^2| \leq \epsilon$; the sequence $q_n/n^2$ converges exactly as stated by the lemma.\slug

@@

\proof The case $n=1$ is obviously true, so let $n>1$. By the fundamental theorem of arithmetic, $n = p_1^{e_1}p_2^{e_2}\cdots p_k^{e_k}$. Only square-free divisors contribute to the sum $\sum_{d\backslash n}\mu(d)$, so we have
$$\sum_{d\backslash n}\mu(d) = \mu(1) + \sum_{1\leq i\leq k} \mu(p_i) + \sum_{1\leq i<j\leq k} \mu(p_ip_j) + \cdots + \mu(p_1\cdots p_k).$$
Since the magnitude of $\mu(d)$ is 1 (whenever $d$ is square-free) and the sign of $\mu(d)$ depends on the number of primes that make up $d$, this can be re-expressed as the alternating sum of binomial coefficients
$$\sum_{d\backslash n}\mu(d) = {k \choose 0} - {k\choose 1} + {k\choose 2} - \cdots + (-1)^k{k\choose k} = (1-1)^k = 0.\noskipslug$$

@@

\proclaim Lemma C.
$$\biggl(\sum_{k\geq 1} {\mu(k) \over k^2}\biggr) \biggl( \sum_{m \geq 1} {1\over m^2}\biggr) = 1.$$
